Testing the thermal buffer of mire groundwater

### 2020-05-09 doc/manuscript/ms.Rmd
polish intro

@@

The dataloggers recorded temperatures for five years in five of the sites, four years in two, and two years in one (Fig. \@ref(fig:fig1)). A visual inspection of the time series shows more homogeneous temperatures in the waterlogged time series, which are almost always encompassed within the dry time series. The bioclimatic variables (Table \@ref(tab:table2)) support this impression. The maximum temperatures of the warmest months are usually higher in the dry measuring points, by more than 6 ºC in some sites, both at low (El Molinucu, La Malva) and high (La Recoleta) elevations. The opposite is true for the minimums of the coldest months, in which case the temperature is generally colder in the dry points, although the difference is less pronounced than for the maximums. The mean diurnal range of temperatures is wider in the dry sites, as is annual temperature range; and again the effect is larger in the low sites and in La Recoleta. The t-tests supported the original hypotheses; namely waterlogged measuring points had (a) smaller diurnal fluctuactions (t = `r ttests %>% filter(Trait == "Mean.Diurnal.Range") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Mean.Diurnal.Range") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Mean.Diurnal.Range") %>% pull(t.estimate) %>% round(2)` ºC); (b) lower maximums (t = `r ttests %>% filter(Trait == "Max.Temperature.of.Warmest.Month") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Max.Temperature.of.Warmest.Month") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Max.Temperature.of.Warmest.Month") %>% pull(t.estimate) %>% round(2)` ºC); (c) higher minimums (t = `r ttests %>% filter(Trait == "Min.Temperature.of.Coldest.Month") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Min.Temperature.of.Coldest.Month") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Min.Temperature.of.Coldest.Month") %>% pull(t.estimate) %>% round(2)` ºC), and (d) smaller annual fluctuations (t = `r ttests %>% filter(Trait == "Temperature.Annual.Range") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Temperature.Annual.Range") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Temperature.Annual.Range") %>% pull(t.estimate) %>% round(2)` ºC).

@@

@@

```{r fig1, echo = FALSE, fig.pos = "H", fig.cap = "Hourly soil temperature records at the mire sites. The blue series was recorded within the mire, in a waterloged area. The red series was recorded in a neighbouring dry area. Dataloggers were buried at 5 cm depth.", fig.align = "center", fig.width=3, fig.height=8}
plot1
```

>>>>

Climate change [@RN3586] affects global biodiversity from drylands [@RN4672] to forests [@RN4673] and oceans [@RN4674]. Ecosystems can be especially vulnerable when they meet certain criteria [@RN4675]: (a) preponderance of species that evolved under a cold climate; (b) low productivity due to nutrient limitation, making them sensitive to increased nutrient cycling caused by warming [@RN4676]; and (c) scatterred distribution pattern, which limits species dispersal and migration [@RN4677]. These characteristics are found in temperate peatlands such as mires, i.e. permanent semi-terrestrial wetlands where soils remain waterlogged but not inundated during most of the year [@RN3161]. Mires are priority habitats for conservation that harbour high numbers of endangered species [@RN3122] and support highly-adapted floras in spatially-reduced areas [@RN2960]. Habitat distribution models have predicted a loss of mire surface as a consequence of ongoing climate change [@RN2937].

Mires are however azonal habitats whose existence depends on local soil properties [@RN3328]: by definition, they are areas with waterlogged soils. Grounwater can produce a buffer effect on soil temperature, keeping soils warmer than the air during cold periods, and vice versa [@RN3344; @RN3201]. Root-zone temperature is a major determinant of plant ecophysiology [@RN3024], so the groundwater buffer effect is expected to allow mire plants to live along a wider range of air temperatures than they could otherwise. Indeed, mires have a relatively homogeneous flora despite being distributed from Mediterranean to Boreal biomes, and from low valleys to the alpine belt [@RN4678]. Recently, soil temperature measurements have become available for mires of North America [@RN3204], Western Europe [@RN2356] and Central Europe [@RN4675]. Their comparison with air temperatures derived from models has shown that mire soils are indeed warmer in winter and cooler in summer, thus giving support to the existence of the groundwater thermal buffer [@RN2356; @RN4675]. Furthermore, the effect has been linked to the composition of mire flora and fauna [@RN4675; @RN4679], the growth rings of mire trees [@RN3060] and the role of mires as glacial refugia [@RN2513; @RN4680].

It is evident that the groundwater buffer effect will play a determinant role in the response of mire habitats to climate change. As is the case for all azonal habitats, locally measured temperatures are essential to understand this response. Available data recorded temperatures only on the waterlogged soils, using model-derived air temperatures for comparison [@RN2356; @RN4675; @RN4679]; or recorded temperatures in both wet and dry spots but for less than a year [@RN3204; @RN3060]. This article provides the first measurement of the thermal buffer against surrounding non-mire areas, based on soil temperatures recorded during a period of five years. These measures are used to test the hypotheses that, when compared to adjacent dry soils, waterlogged mire soil are (i) warmer in winter and (ii) colder in summer; and have less thermal amplitude in (iii) daily and (iv) annual scales.

@@

The dataloggers recorded temperatures for five years in five of the sites, four years in two, and two years in one (Fig. \@ref(fig:fig1)). A visual inspection of the time series shows more homogeneous temperatures in the waterlogged time series, which are almost always encompassed within the dry time series. The bioclimatic variables (Table \@ref(tab:table2)) support this impression. The maximum temperatures of the warmest months are usually higher in the dry measuring points, by more than 6 ºC in some sites, both at low (El Molinucu, La Malva) and high (La Recoleta) elevations. The opposite is true for the minimums of the coldest months, in which case the temperature is generally colder in the dry points, although the difference is less pronounced than for the maximums. The mean diurnal range of temperatures is wider in the dry sites, as is annual temperature range; and again the effect is larger in the low sites and in La Recoleta. The t-tests supported the original hypotheses (Fig. \@ref(fig:fig2)); namely waterlogged measuring points had (a) smaller diurnal fluctuactions (t = `r ttests %>% filter(Trait == "Mean.Diurnal.Range") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Mean.Diurnal.Range") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Mean.Diurnal.Range") %>% pull(t.estimate) %>% round(2)` ºC); (b) lower maximums (t = `r ttests %>% filter(Trait == "Max.Temperature.of.Warmest.Month") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Max.Temperature.of.Warmest.Month") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Max.Temperature.of.Warmest.Month") %>% pull(t.estimate) %>% round(2)` ºC); (c) higher minimums (t = `r ttests %>% filter(Trait == "Min.Temperature.of.Coldest.Month") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Min.Temperature.of.Coldest.Month") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Min.Temperature.of.Coldest.Month") %>% pull(t.estimate) %>% round(2)` ºC), and (d) smaller annual fluctuations (t = `r ttests %>% filter(Trait == "Temperature.Annual.Range") %>% pull(t.statistic) %>% round(2)`, p = `r ttests %>% filter(Trait == "Temperature.Annual.Range") %>% pull(t.p.value) %>% round(3)`, effect size = `r ttests %>% filter(Trait == "Temperature.Annual.Range") %>% pull(t.estimate) %>% round(2)` ºC).

@@

Less effect in winter could be because this is soil (snow, etc)

@@

```{r fig1, echo = FALSE, fig.pos = "H", fig.cap = "Hourly soil temperature records at the mire sites. The blue series was recorded within the mire, in a waterloged area. The red series was recorded in a neighbouring dry area. Dataloggers were buried at 5 cm depth.", fig.align = "center", fig.width=7, fig.height=6}
plot1
```

```{r fig2, echo = FALSE, fig.pos = "H", fig.cap = "Average bioclimatic variables in the dry and waterlogged sites. The bars represent the mean value, and the bars the standard error of 8 records", fig.align = "center", fig.width=7, fig.height=3}
plot2
```
